Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-08-22 analysis/paper/paper-doc.qmd

Sample size can impact on accuracy of estimated CV that reflects the corresponding population size. [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In archeological research, sample size for CV values varies from n=5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements on small samples. @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e. n\<25). Statistical research has results in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to show uncertainty around our CV estimates @kelley2007sample; @albatineh2014con fidence].

@@

PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission.Figure 5 shows positive relationship between attributes. Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length. Body length and maximum length contribute the most to the shape of stemmed points and they are most correlated to each other. Other width and tang related attributes are close to each other but a bit far from body length and maximum length. Stem width contributes the least to the overall shape. We can assume that

>>>>

Sample size can impact on accuracy of estimated CV that reflects the corresponding population size. [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In archeological research, sample size for CV values varies from n=5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements on small samples. @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e. n\<25). Statistical research has results in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to show uncertainty around our CV estimates [@kelley2007sample; @albatineh2014confidence].

@@

PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission. @fig-five shows positive relationship between attributes. Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length. Body length and maximum length contribute the most to the shape of stemmed points and they are most correlated to each other. Other width and tang related attributes are close to each other but a bit far from body length and maximum length. Stem width contributes the least to the overall shape. We can assume that the body length and maximum length were prioritized when the stemmed points were transmitted.

![PCA analysis for stemmed points. The direction of the arrow shows the relationship between attributes. The length and color of the arrow indicate the strength of contribution to the stemmed point shape.](images/007-PCA-1.png){#fig-five width="284"}

4.  Variation Over Time

To understand the temporal patterns in the modes of cultural transmission, we applied Korea Late Paleolithic chronological phases into our samples and computed CV and PCA. Among the three phases of the chronology, we excluded Phase 1 for the analysis because there are only two stemmed points from Yonghodong and Sachang sites to compute. However, this does not mean that there were only two stemmed points for the Phase 1 because we excluded broken pieces from our sample to conduct the morphological analysis.

After excluding the Phase 1 stemmed points, overall CV values become lower (@fig-six). Body length of Phase 2, maximum length, mid width, and tang width are under 25, which indicates as a result of indirect bias, while only maximum length are under 25 in Figure 4. On the other hand, Body length of Phase 3, stem length, stem width, and tang length are relatively higher then 25 and mostly over 30 indicating guided variation.

Compared to Phase 2, except for maximum width and tang width, the other CV values of Phase 3 tend to be higher. In addition, Phase 3 tends to have wider confidence intervals. Stem width and tang length values are relatively steady over time. Our results show that stemmed points from Phase 2 are less varied than Phase 3 as well as Phase 1.

![CV values for the second and third chronological phases in the Korean Late Paleolithic period. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.](images/004-cv-phases-01.png){#fig-six width="346"}

Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective (@fig-seven). However, there are no notable changes in correlations over time.

![PCA analysis for the second and third chronological phases in the Korean Late Paleolithic period. The direction of the arrow shows the relationship between attributes. The length and color of the arrow indicate the strength of contribution to the stemmed point shape.](images/007-PCA.png){#fig-seven width="534"}

We explored assemblages that contain multiple stemmed points to examine the temporal patterns with the premise that stemmed points from one assemblage were made by the same group of people. There are four assemblages, three assemblages from the Suyanggae site (SYG) and one from the Yongsandong site (YS) that contain seven or more stemmed points. SYG1_2 and SYG6_2 are from Phase 2 and SYG1_3 and YS_3 are from Phase 3. Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-eight shows that CV values for the four assemblages are mostly below 25, which represent transmissions through indirect bias. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest range of confidence intervals. Compared to the Phase 2, CV values of Phase 3 are lower, which is different from @fig-six results of body length and maximum length from all assemblages. It is assumed that those attributes became more standardized within the same assemblages. Compared to Suyanggae assemblages, Yongsandong has smaller CV values.

![CV values of four attributes from assemblages that have 7 or more stemmed points. Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.](images/006-cv-four-assemblage.png){#fig-eight width="386"}

5.  Variation Between sites

We examined the CV values per site to address our research question about the cultural transmission over space. We excluded assemblages that contain a single point. @fig-nine shows that Sachang site has the highest CV values with the widest confidence level. Sachang is part of Phase 1 and its CV values explain why the CV of all time is higher than Phase 2. We excluded sites that have only one stemmed point for the analysis. Two assemblages from the Suyanggae site have relatively stable CV for attributes, which indicate highly standardized shapes of stemmed points. CV values of body length is the lowest while stem width the highest among all assemblages. Overall, there are no clear patterns between the sites.

![CV values per site that has more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.](images/004-cv-sites-01.png){#fig-nine width="519"}

6.  Variation Between Raw Materials

We examined the relationship between CV values and raw materials with a premise that the shape of stemmed points are highly dependent on raw materials. We exclude raw materials that were used for single point. @fig-ten shows that quartzite, the raw materials for Sachang stemmed points, has the highest CV value with a wide range of conference intervals. Quartz has high variations of CV among the attributes. On the other hand, shale has the lower CV, which is very stable throughout the attributes.

![CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.](images/00000-cv-raw-materials.png){#fig-ten width="526"}

# Discussion {.unnumbered}

# References {.unnumbered}
